---
title: "PS270 Final Project"
author: Rares-Matei Puiu
description: "Do Romanian working-class voters feel left behind?"
output:
  distill::distill_article:
    self_contained: false
---

```{r, echo=F}
rm(list=ls())
library(tidyverse)
```


# **Milestone 2: Dataset and Research Question**

I am currently refining the precise focus of this project, but at this stage I plan to use a dataset derived from an **AtlasIntel** national poll fielded November 9–12, 2024 that reports Călin Georgescu’s vote intention as percentages within demographic subgroups (including age, income, and education), alongside a benchmark row recording his official first-round vote share (22.94%). I made the csv file myself so bare with it as I further develop and refine it for the final deadline.

While I initially considered education as the primary explanatory variable, I now believe income level provides a clearer theoretical basis for analysis. My tentative research question is: **How is Georgescu’s reported support in the poll related to respondents’ income level?** I hypothesize that Georgescu’s support will be higher among lower-income respondents and lower among higher-income respondents. The theoretical logic is that anti-establishment candidates often mobilize voters who perceive themselves as economically insecure or excluded from the benefits of the existing political and economic order, whereas higher-income voters may have stronger incentives to prefer stability and institutional continuity.

My explanatory variable is **income level**, measured using the ordered income categories reported in the poll (e.g., <3000 RON, 3000–5000 RON, 5000–7000 RON, etc.), which provides observable variation across multiple brackets. My outcome variable is **Georgescu’s vote intention percentage**, measured as georgescu_pct within each income subgroup. Evidence supporting my hypothesis would be a statistically significant negative relationship between ordered income level and Georgescu’s support in a regression model (optionally weighted by subgroup sample size), indicating that support declines as income increases. Evidence against the hypothesis would be a relationship near zero or positive, suggesting that income does not systematically structure support for Georgescu in this poll.




```{r, echo=F}
library(knitr)



poll <- read_csv("data/georgescu.csv", show_col_types = FALSE)

poll |>
  select(sample_n, segment_type, segment, georgescu_pct) |>
  tail(5) |>
  rename(
    "Subgroup Sample" = sample_n,
    "Category Type" = segment_type,
    "Category" = segment,
    "Georgescu Support (%)" = georgescu_pct
  ) |>
  kable(digits = 1)
```
_Source: AtlasIntel national poll (Nov 9–12, 2024)._